---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phsopendata

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/phsopendata)](https://CRAN.R-project.org/package=phsopendata)
[![GitHub release (latest by date)](https://img.shields.io/github/v/release/Public-Health-Scotland/phsopendata)](https://github.com/Public-Health-Scotland/phsopendata/releases/latest)
[![R-CMD-check](https://github.com/Public-Health-Scotland/phsopendata/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Public-Health-Scotland/phsopendata/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Public-Health-Scotland/phsopendata/branch/master/graph/badge.svg)](https://app.codecov.io/gh/Public-Health-Scotland/phsopendata?branch=master)
<!-- badges: end -->

`phsopendata` contains functions to interact with open data from the [Scottish Health and Social Care Open Data platform](https://www.opendata.nhs.scot/) via the CKAN API. 

- `get_resource()` extracts a single resource from an open dataset by resource id
- `get_latest_resource()` extracts the most recent resource from applicable datasets, by dataset name
- `get_dataset()` extracts multiple resources from an open dataset by dataset name
- `list_datasets()` returns the names of all available datasets 
- `list_resources()` returns information on all resources within an open dataset by dataset name


`phsopendata` can be used on both Posit Workbench and desktop versions of RStudio.

## Installation

```{r, eval = FALSE}
# The easiest way to get phsopendata is to install from CRAN:
install.packages("phsopendata")
```

### Development version

To get a bug fix or to use a feature from the development version, you can install 
the development version of phsopendata from GitHub.

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("Public-Health-Scotland/phsopendata")
```

## Examples

### Downloading a data table with `get_resource()`

To extract a specific resource, you will need its unique identifier - resource id. This can be found in the dataset metadata, the URL of a resource's page on https://www.opendata.nhs.scot/, or extracted using `list_resources()`.

```{r example resource, eval = FALSE}
library(phsopendata)

# define a resource ID
res_id <- "a794d603-95ab-4309-8c92-b48970478c14"

# download the data from the CKAN database
data <- get_resource(res_id = "a794d603-95ab-4309-8c92-b48970478c14")
```

### Querying/filtering data with `get_resource()`
You can define a row limit with the `rows` argument to get the first *N* rows of a table.
```{r example row, eval = FALSE}
# get first 100 rows
get_resource(
  res_id = "a794d603-95ab-4309-8c92-b48970478c14",
  rows = 100
)
```

You can use `col_select` and `row_filters` to query the data server-side (i.e., the data is filtered before it is downloaded to your machine).
```{r example query, eval = FALSE}
# get first 100 rows
get_resource(
  res_id = "a794d603-95ab-4309-8c92-b48970478c14",
  col_select = c("GPPracticeName", "TelephoneNumber"),
  row_filters = list(
    HB = "S08000017",
    Dispensing = "Y"
  )
)
```

### Downloading multiple tables with `get_dataset()`

To extract all resources from a dataset, you will need to use the *dataset name*. Note that this will differ from the *dataset title* that displays on the website. This can be found using `list_datasets()`, or taken from the dataset URL. 

In this example, we are downloading GP Practice Population Demographics from: [opendata.nhs.scot/dataset/*gp-practice-populations*](https://www.opendata.nhs.scot/dataset/gp-practice-populations), so the dataset name will be gp-practice-populations.


```{r example dataset, eval = FALSE}
# if max_resources is not set, all resources will be returned by default.
# Here we pull 10 rows from the first 2 resources only
get_dataset("gp-practice-populations", max_resources = 2, rows = 10)
```

## Contributing to phsopendata

At present, this package is maintained by [James Hayes](https://github.com/Moohan).

If you have requests or suggestions for additional functionality, please contact the package maintainer and/or the [PHS Open Data team](mailto:phs.opendata@phs.scot).

If you would like to share examples of how you work with open data, you can also do so in the [Open Data repository](https://github.com/Public-Health-Scotland/Open-Data), where example scripts and resources are collated.
