---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phsopendata

<!-- badges: start -->
<!-- badges: end -->

`phsopendata` contains functions to interact with open data from the [Scottish Health and Social Care Open Data platform](https://www.opendata.nhs.scot/) via the CKAN API. 

- `get-resource` extracts a single resource from an open dataset by resource id
- `get-dataset` extracts multiple resources from an open dataset by dataset name

For extracting metadata and search functionality, we recommend using the [ckanr package](https://docs.ropensci.org/ckanr/). 

`phsopendata` can be used on both the [server](https://rstudio.nhsnss.scot.nhs.uk/) and desktop versions of RStudio. However, depending on firewall settings, proxy use may need to be configured with `use_proxy()`.

## Installation

To install `phsopendata`, the package `remotes` is required, and can be installed with `install.packages("remotes")`.

You can then install `phsopendata` on RStudio server from GitHub with:

```{r gh-installation, eval = FALSE}
remotes::install_github("Public-Health-Scotland/phsopendata",
  upgrade = "never"
)
```

## Examples

### get_resource

To extract a specific resource, you will need it's unique identifier - resource id. This can be found in the dataset metadata by visiting the website, or extracted using `ckanr::package_show`.

```{r example resource, eval = FALSE}
library(phsopendata)

#by default the full resource is returned
opendata_get_resource(res_id = "a794d603-95ab-4309-8c92-b48970478c14")

#but you can set the number of rows to return
opendata_get_resource(res_id = "a794d603-95ab-4309-8c92-b48970478c14", rows = 10)
```

### get_dataset

To extract all resources from a dataset, you will need to use the *dataset name*. Note that this will differ from the *dataset title* that displays on the website. This can be found in the dataset metadata extracted using `ckanr::package_show`, or taken from the dataset URL. 

In this example, we are downloading GP Practice Population Demographics from: opendata.nhs.scot/dataset/gp-practice-populations, so the dataset name will be gp-practice-populations.


```{r example dataset, eval = FALSE}
#if max_resources is not set, all resources will be returned by default. Here we pull 10 rows from the first 2 #resources only
opendata_get_dataset("gp-practice-populations", max_resources = 2, rows = 10)
```

## Contributing to phsopendata

At present, this package is maintained by [Csilla Scharle](https://github.com/csillasch).

If you have requests or suggestions for additional functionality, please contact the package maintainer and/or the [PHS Open Data team](phs.opendata@phs.scot).

If you would like to share examples of how you work with open data, you can also do so in the [Open Data repository](https://github.com/Public-Health-Scotland/Open-Data), where example scripts and resources are collated.
