---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phsopendata

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/phsopendata)](https://CRAN.R-project.org/package=phsopendata)
[![GitHub release (latest by date)](https://img.shields.io/github/v/release/Public-Health-Scotland/phsopendata)](https://github.com/Public-Health-Scotland/phsopendata/releases/latest)
[![PHS package checks](https://github.com/Public-Health-Scotland/phsopendata/actions/workflows/phs_checks.yaml/badge.svg)](https://github.com/Public-Health-Scotland/phsopendata/actions/workflows/phs_checks.yaml)
[![Codecov test coverage](https://codecov.io/gh/Public-Health-Scotland/phsopendata/branch/master/graph/badge.svg)](https://app.codecov.io/gh/Public-Health-Scotland/phsopendata?branch=master)

<!-- badges: end -->

`phsopendata` provides helper functions for discovering and downloading data from the [Scottish Health and Social Care Open Data platform](https://www.opendata.nhs.scot/) using the CKAN API.

It can be used to:

- search for datasets and resources;
- download a single resource by resource ID;
- download multiple resources from a dataset;
- filter rows and select columns before downloading data;
- retrieve the latest resource from datasets that publish new resources over time;
- run SQL queries against Open Data resources.

See the [package website](https://public-health-scotland.github.io/phsopendata/) for full documentation.

## Installation

```{r, eval = FALSE}
# The easiest way to get phsopendata is to install from CRAN:
install.packages("phsopendata")
```

### Development version

To get a bug fix or to use a feature from the development version, you can install the development version of phsopendata from GitHub.

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("Public-Health-Scotland/phsopendata")
```

## Quick start

### Find resources

To download data, you will usually need either a `dataset_name` or a `resource_id`.

These can be found in the dataset metadata, in the URL of a dataset or resource page on <https://www.opendata.nhs.scot/>, or by searching with `list_resources()`. `list_datasets()` is still available for compatibility, but `list_resources()` is recommended for new code.

```{r example_search, eval = FALSE}
library(phsopendata)

resources <- list_resources(
  dataset_contains = "gp practice",
  resource_contains = "contact"
)

all_datasets <- list_datasets()
```

### Download a resource

```{r example_resource, eval = FALSE}
# define a resource ID
res_id <- "a794d603-95ab-4309-8c92-b48970478c14"

# download the data
open_data <- get_resource(res_id)
```

### Filter rows and select columns

You can define a row limit with the `rows` argument to get the first *N* rows of a table.

```{r example_row, eval = FALSE}
# get first 100 rows
open_data <- get_resource(
  res_id = res_id,
  rows = 100
)
```

You can use `col_select` and `row_filters` to query the data server-side (i.e., the data is filtered before it is downloaded to your machine).

```{r example_query, eval = FALSE}
# select columns and filter rows before downloading
open_data <- get_resource(
  res_id = res_id,
  col_select = c("GPPracticeName", "TelephoneNumber"),
  row_filters = list(
    HB = "S08000017",
    Dispensing = "Y"
  )
)
```

### Download a dataset

In this example, we are downloading GP Practice Population Demographics from: [opendata.nhs.scot/dataset/*gp-practice-populations*](https://www.opendata.nhs.scot/dataset/gp-practice-populations), so the dataset name will be gp-practice-populations.

```{r example_dataset, eval = FALSE}
# if max_resources is not set, all resources will be returned by default.
# Here we pull 10 rows from the first 2 resources only
practice_pops <- get_dataset("gp-practice-populations", max_resources = 2, rows = 10)
```

### Download the latest resource from a dataset

Some datasets publish new resources over time rather than replacing an existing resource. For these datasets, you can use `get_latest_resource()`.

```{r example_latest, eval = FALSE}
latest_contacts <- get_latest_resource(
  dataset_name = "gp-practice-contact-details-and-list-sizes",
  col_select = c("PracticeCode", "PracticeName", "Postcode", "Dispensing")
)

### Query using SQL

For more flexible server-side queries, use `get_resource_sql()`. SQL queries can return a maximum of 32,000 rows.

```{r example_sql, eval = FALSE}
cancelled_ops <- get_resource_sql(r"[
SELECT
    "Hospital",
    "Month",
    "TotalCancelled",
    "TotalOperations"
FROM
    "bcc860a4-49f4-4232-a76b-f559cf6eb885"
WHERE
    "Hospital" = 'D102H'
]")
```

## Function overview

- `list_resources()` searches available datasets and resources on the Open Data platform.
- `get_resource()` downloads a single resource by resource ID, with optional row filtering and column selection.
- `get_dataset()` downloads multiple resources from a dataset by dataset name.
- `get_latest_resource()` downloads the most recent resource from datasets that publish new resources over time.
- `get_resource_sql()` runs a SQL query against one or more Open Data resources.
- `get_dataset_additional_info()` returns summary information about a dataset, including the number of resources and the latest update date.
- `list_datasets()` returns dataset names only. It is retained for compatibility but has been superseded by `list_resources()`.

## Contributing

This package is maintained by Public Health Scotland.

For requests, bug reports or suggestions, please use the GitHub issues page or contact the [PHS Open Data team](mailto:phs.opendata@phs.scot).

If you would like to share examples of how you work with open data, you can also do so in the [Open Data repository](https://github.com/Public-Health-Scotland/Open-Data), where example scripts and resources are collated.
